The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2019-10-30 categorical-zero-one.tex
Add stuff about Kl(H)

\begin{lemma}[The determinism lemma]
	\label{determinismlemma}
	Suppose that we have $p: A \to X$ and deterministic $s : X \to T$. If the joint
	\begin{equation}
		\label{determinismlemma_joint}
		\tikzfig{determinismlemma_joint}	
	\end{equation}
	displays the conditional independence $X \perp T \mid\mid A$, then the composite $sp: A \to T$ is deterministic.
\end{lemma}\cite
This result can be thought of as a partial converse to the condition which characterizes positivity of a Markov category~\cite[Definition~11.22]{markov_cats}. Here and in the following, our notation ``$s$'' indicates that we think of this morphism as a statistic~\cite[Definition~14.2]{markov_cats}.
\begin{proof}
	Since the marginals of \Cref{determinismlemma_joint} are exactly $p$ and $sp$, the conditional independence assumption is
	\[
		\tikzfig{determinismlemma1}
	\]
	Using this equation together with the assumption that $s$ is deterministic, we then compute
	\[
		\tikzfig{determinismlemma2}
	\]
	which is what was to be shown.
\end{proof}

@@

By interpreting our results in other Markov categories, we obtain other nontrivial (although hardly particularly deep) statements:

\todo[inline]{Describe Kleisli cat of the hyperspace monad, or directly whatever precise category of relations the following takes place in}

\begin{corollary}
    Let $(X_j)_{j \in J}$ be a family of topological spaces, $Y$ a Hausdorff space, and let $f: \prod_j X_j \to Y$ be a continuous function
    which is independent of any finite prefix of the input.
    Then the map $f$ is constant.
\end{corollary}
\begin{corollary}
	Let $X$ be a topological space, $Y$ a Hausdorff space, $J$ an infinite set, and let $f: \prod_{j\in J} X \to Y$ be a continuous function which is invariant under any finite permutation of the inputs.
    Then $f$ is constant.
\end{corollary}
\begin{proof}[Proof of both]
	These results follow from \Cref{thm:kolmog,thm:hewsav}, applied to the Kleisli category $\mathrm{Kl}(H)$ of the hyperspace monad $H$,
	using the morphisms $p : I \to X_J$ corresponding to the trivial closed subset $X_J \subseteq X_J$.
    The conclusion is that the closure of the image, call it $A$, has the following property: $\overline{\{(a,a) \mid a \in A\}} = A \times A$ as subsets of $Y \times Y$.
    But since the diagonal is closed in $Y$, this clearly implies that $A$ must be a singleton.
\end{proof}

\todo[inline]{I agree with the speculation, but it's probably a bit too vague for the paper in its present state, so I've removed it}

>>>>

\begin{lemma}[The determinism lemma]
	\label{determinismlemma}
	Suppose that we have $p: A \to X$ and deterministic $s : X \to T$. If the joint
	\begin{equation}
		\label{determinismlemma_joint}
		\tikzfig{determinismlemma_joint}	
	\end{equation}
	displays the conditional independence $X \perp T \mid\mid A$, then the composite $sp: A \to T$ is deterministic.
\end{lemma}
This result can be thought of as a partial converse to the condition which characterizes positivity of a Markov category~\cite[Definition~11.22]{markov_cats}. Here and in the following, our notation ``$s$'' indicates that we think of this morphism as a statistic~\cite[Definition~14.2]{markov_cats}.
\begin{proof}
	Since the marginals of \Cref{determinismlemma_joint} are exactly $p$ and $sp$, the conditional independence assumption is
	\[
		\tikzfig{determinismlemma1}
	\]
	Using this equation together with the assumption that $s$ is deterministic, we then compute
	\[
		\tikzfig{determinismlemma2}
	\]
	which is what was to be shown.
\end{proof}

@@

\todo[inline]{E: Not sure these examples are that great.. it seems like we do an inordinate amount of work to get the infinite tensor products, compared to the results. What do you think?}

By interpreting these results in different Markov categories, we can obtain other results.
As a simple example, we can consider the following Markov category:

\begin{definition}
    Consider the hyperspace monad on $\Top$ defined in \cite[Definition~2.4]{hyperspace}.
    The Kleisli category associated to this monad is denoted $\Kl(H)$.
    Using \cite[Corollary~2.53]{hyperspace} and \cite[Proposition~3.1]{markov_cats}, this is a Markov category.
    Concretely, this category has the following structure:
    \begin{enumerate}
        \item The objects are topological spaces.
        \item A map $f: X \to Y$ consists of a continuous function $X \to HY$, where $HY$ denotes the set of closed subsets, equipped with the topology generated by subbasic opens of the form $\{V \subset Y \text{ closed} \mid V \cap U \neq \emptyset\}$ for $U \subset Y$ open.
        \item Composition is defined as in $\SetMulti$, by the closure of the union.
        \item The monoidal product is the Cartesian product of topological spaces, and $(f \tensor g)(x_1,x_2) = f(x_1) \times g(x_2)$.
        \item The comultiplication is defined by $\cop{X}(x) = \cl(\{(x,x)\})$
    \end{enumerate}
\end{definition}
\begin{proposition}
    $\Kl(H)$ has Kolmogorov products of any cardinality, given simply by the ordinary infinite product of topological spaces in the product topology.
\end{proposition}
\begin{proof}
    First we observe that $H(\prod_{j\in J} X_j)$ is the cofiltered limit of the spaces
    $H(\prod_{j\in F} X_j)$.
    To see this, note that the marginalizations induce a continuous map
    \[\alpha: H(\prod_{j\in J} X_j) \to \lim_{F \subset J} H(\prod_{j \in J} X_j)\]
    This is an injection - given distinct closed subsets $V, W \subset \prod X_j$, and let $x \in V \setminus W$.
    Then let $U \ni x$ be an open neighborhood of $x$ not intersecting $W$.
    Since $U$ is open in the product topology, we can find some finite subset $F \subset J$ so that 
    \[y = (y_j) \in U \Leftrightarrow y_j \in \pi_j(U) \text{ for $j\in F$}\]
    Then $\pi_F(x) \in \pi_F(U)$, which is disjoint from $\pi_F(W)$. Hence $\alpha(V) \neq \alpha(W)$.

    It is also a surjection - given a family $(V_F) \in \lim H(\prod_{j \in J} X_j)$, simply set
    \[V = \{x \in \prod_{j\in J} X_j \mid  \pi_F(x) \in V_F \forall F\}\]
    This is an intersection of closed subsets, hence closed, and it is clearly sent to $(V_F)$.

    Finally, we must verify that it's open.
    It suffices to show that it carries subbasis open sets $A_U = \{V \subset \prod X_j \text{ closed} \mid V \cap U \neq \emptyset\}$ to open subsets of the limit.
    Again, given such a $U$, we find a finite $F$ so that
    \[y = (y_j) \in U \Leftrightarrow y_j \in \pi_j(U) \text{ for $j\in F$}\]
    then \[\alpha(A_U) = \pi_F\inv(\{V \subset \prod_{j \in F} X_F \text{ closed} \mid V \cap \pi_F(U) \neq \emptyset\})\]
    and this is open in the limit.

    It follows that $\prod_j X_j$ is an infinite tensor product in $\Kl(H)$ - the preservation condition is easily verified.
    The Kolmogorov product condition is similarly immediate to verify.
\end{proof}
\begin{corollary}
    Let $(X_j)_{j \in J}$ be a family of topological spaces, $Y$ a Hausdorff space, and let $f: \prod_j X_j \to Y$ be a continuous function
    which is independent of any finite prefix of the input.
    Then the map $f$ is constant.
\end{corollary}
\begin{corollary}
	Let $X$ be a topological space, $Y$ a Hausdorff space, $J$ an infinite set, and let $f: \prod_{j\in J} X \to Y$ be a continuous function which is invariant under any finite permutation of the inputs.
    Then $f$ is constant.
\end{corollary}
\begin{proof}[Proof of both]
	These results follow from \Cref{thm:kolmog,thm:hewsav} appliedto $\Kl(H)$.
	using the morphisms $p : I \to X_J$ corresponding to the trivial closed subset $X_J \subseteq X_J$.
    The conclusion is that the closure of the image, call it $A$, has the following property: $\cl(\{(a,a) \mid a \in A\}) = A \times A$ as subsets of $Y \times Y$.
    But since the diagonal is closed in $Y$, this clearly implies that $A$ must be a singleton.
\end{proof}
